`conmat`: generate synthetic contact matrices for a given age-stratified population

### 2025-12-18 paper/paper.qmd

---
title: '`conmat`: generate synthetic contact matrices for a given age-stratified population'
format: 
  html: 
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags:
  - epidemiology
  - R
  - infectious disease
authors:
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    corresponding: true
    affiliation: "1"
  - name: Chitra Saraswati
    orcid: 0000-0002-8159-0414
    affiliation: "1"
  - name: Aarathy Babu
    orcid: 0000-0002-6982-5989
    affiliation: "1"
  - name: Michael Lydeamore
    orcid: 0000-0001-6515-827X
    affiliation: "3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1,2"
affiliations:
  - index: 1
    name: "The Kids Research Institute Australia, WA, Australia"
  - index: 2
    name:  "University of Western Australia, WA, Australia"
  - index: 3
    name: "Department of Econometrics and Business Statistics, Monash University, VIC, Australia"
date: today
bibliography: paper.bib
---

@@

There are currently few options for a user to access synthetic contact matrices [@socialmixr; @prem2017]. Existing code to generate synthetic contact matrices from @prem2017 are not designed for replicability, are restricted to select countries, and provide no sub-national demographic estimates.

The `conmat` package exposes model fitting and prediction separately to the user. Users can fit a model based on a contact survey such as POLYMOD [@mossong2008], then predict from this model to their own demographic data. This means users can generate synthetic contact matrices for any region, with any contact survey.

@@

Infectious diseases like influenza and COVID-19 spread via social contact. If we can understand patterns of contact---which individuals are more likely be in contact with each other---then we will be able to create models of how disease spreads. Epidemiologists and public policy makers can use these models to make decisions to keep a population safe and healthy.

Empirical estimates of social contact are provided by social contact surveys. These provide samples of the frequency and type of social contact across different settings (home, work, school, other).

A prominent contact survey is the POLYMOD study by @mossong2008, which surveyed 8 European countries: Belgium, Germany, Finland, Great Britain, Italy, Luxembourg, The Netherlands, and Poland [@mossong2008].

These social contact surveys can be projected on to a given demographic structure to produce estimated daily contact rates between age groups. These are known as 'synthetic' contact matrices. A widely used approach by @prem2017 [@prem2021] produced synthetic contact matrices for 177 countries at 'urban' and 'rural' levels for each country.

However, there were major limitations with the methods in @prem2021. First, not all countries were included in their analyses. Second, the contact matrices only covered broad population groups within entire countries. This presents challenges for decision makers who are often working at a sub-national geographical scale, with differing demographic structure in different sub-populations. Third, the code provided by Prem et al. was not designed for replicability and easy modification with user-defined inputs.

The `conmat` package was developed to fill the specific need of creating contact matrices for arbitrary age categories and populations (as shown in the below example) to inform infectious diease models. We developed the method primarily to output synthetic contact matrices. We also provided methods to create next generation matrices for modelling.

@@

`conmat` was built to predict at four settings: work, school, home, and other. 
One model is fitted for each setting. 
Each model fitted is a Poisson generalised additive model (GAM) with a log link function, which predicts the count of contacts, with an offset for the log of participants. 
The model has six covariates to explain six key features of the relationship between ages, 
and two optional covariates for attendance at school or work.
The two optional covariates are included depending on which setting the model is fitted for.

@@

The covariate capture typical features of inter-person contact, where individuals primarily interact with people of similar age (the diagonals of the matrix), and with grandparents and/or children (the so-called 'wings' of the matrix). The key features of the relationship between the age groups, represented by spline transformations of the six covariates, are displayed in Figure \ref{fig-show-partial-plots} for the home setting. 
```{r}
#| label: partial-plots-create
#| echo: false

@@

The `conmat` software provides a flexible interface to generating synthetic contact matrices using population data and contact surveys. These contact matrices can then be used in infectious disease modelling and surveillance.

@@

The interface to the model formula in `conmat` is fixed; users cannot change the covariates of the model. This means if there is an unusual structure in their contact data it might not be accurately captured by `conmat`. This fixed formula was a design decision made to focus on the key feature of `conmat`: using only age population data to predict a contact matrix.

Public health decisions are often based on age specific information, which means the more accurate your age specific models are, the better those decisions are likely to be. This is the first piece of software that will provide appropriate contact matrices for a population, which means more accurate models of disease.

>>>>

---
title: '`conmat`: generate synthetic contact matrices for a given age-stratified population'
format: 
  wordcount-html: 
    wordcounter-banner: true
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
params:
  wordcount: |
    <strong>{{< words-sum body-note-ref >}} total words</strong>: {{< words-body >}} in the body • {{< words-ref >}} in the references • {{< words-note >}} in the notes
tags:
  - epidemiology
  - R
  - infectious disease
authors:
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    corresponding: true
    affiliation: "1"
  - name: Chitra Saraswati
    orcid: 0000-0002-8159-0414
    affiliation: "1"
  - name: Aarathy Babu
    orcid: 0000-0002-6982-5989
    affiliation: "1"
  - name: Michael Lydeamore
    orcid: 0000-0001-6515-827X
    affiliation: "3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1,2"
affiliations:
  - index: 1
    name: "The Kids Research Institute Australia, WA, Australia"
  - index: 2
    name:  "University of Western Australia, WA, Australia"
  - index: 3
    name: "Department of Econometrics and Business Statistics, Monash University, VIC, Australia"
date: today
bibliography: paper.bib
---

@@

<!-- There are currently few options for a user to access synthetic contact matrices [@socialmixr; @prem2017]. Existing code to generate synthetic contact matrices from @prem2017 are not designed for replicability, are restricted to select countries, and provide no sub-national demographic estimates. -->

The `conmat` package exposes model fitting and prediction separately to the user. Users can fit a model based on a contact survey, then predict from this model to their own demographic data. This means users can generate synthetic contact matrices for any region, with any contact survey.

@@

Infectious diseases like influenza and COVID-19 spread via social contact. If we can understand patterns of contact---which individuals are more likely be in contact with each other---then we will be able to create accurate models of how disease spreads. Epidemiologists and public policy makers can use these models to make decisions to keep a population safe and healthy.

Empirical estimates of social contact are provided by social contact surveys, which provide samples of the frequency and type of social contact across different settings (home, work, school, other). These surveys can be projected onto a given demographic structure, resulting in what is known as 'synthetic' contact matrices. A widely used approach by @prem2017 [@prem2021] produced such matrices for 177 countries at 'urban' and 'rural' levels for each country using data from 8 European Countries by @mossong2008.

However, there were major limitations with the methods in @prem2021. First, not all countries were included in their analyses. Second, the contact matrices only covered broad population groups within entire countries. Third, the code provided by Prem et al. was not designed for replicability and easy modification with user-defined inputs.

The `conmat` package was developed to fill the specific need of creating contact matrices for arbitrary age categories and populations to inform infectious diease models. We developed the method to output synthetic contact matrices. We also provided methods to create next generation matrices.

@@

`conmat` was built to predict at four settings: work, school, home, and other. 
One model is fitted for each setting. 
Each model fitted is a Poisson generalised additive model (GAM) with a log link function, which predicts the count of contacts, with an offset for the log of participants. 
The model has six covariates to explain key features of the relationship between ages, 
and two optional covariates for attendance at school or work.

@@

The model captures typical features of inter-person contact, where individuals primarily interact with people of similar age (the diagonals of the matrix), and with grandparents and/or children (the so-called 'wings' of the matrix). The key features of the relationship between the age groups are displayed in Figure \ref{fig-show-partial-plots} for the home setting. 
```{r}
#| label: partial-plots-create
#| echo: false

@@

The `conmat` software provides a flexible interface to generating synthetic contact matrices using population data and contact surveys. These contact matrices can then be used in infectious disease modelling and surveillance. Public health decisions are based on models using these matrices as input data. This is the first piece of software that can provide context-appropriate contact matrices for a population, which means more accurate models of disease.

@@

<!-- The interface to the model formula in `conmat` is fixed; users cannot change the covariates of the model. This means if there is an unusual structure in their contact data it might not be accurately captured by `conmat`. This fixed formula was a design decision made to focus on the key feature of `conmat`: using only age population data to predict a contact matrix.  -->
